In-silico molecular enrichment and clearance of the human intracranial space

### 2025-01-27 paper/main.tex

% Tracer enrichment predictions
While molecular enrichment of the brain and surrounding CSF spaces
have been extensively studied in animal models, especially in murine
models in connection with studies of the glymphatic system, there are
fewer reports of contrast enrichment in human subjects over 0--24h. We
therefore mainly compare our in-silico predictions with the series of
papers by Eide and colleagues~\cite{ringstad2017glymphatic,
  ringstad2018brain, eide2021sleep, eide2024functional} and Watts et
al~\cite{watts2019measuring}. Ringstad et al.~\cite{ringstad2018brain}
report of tracer enrichment in a centripetal pattern, primarily in
regions near large cerebral surface arteries, peaking in the CSF
spaces between 6 and 24 hours, with parenchymal tracer content still
increasing after 24 hours, but with large individual
differences. Watts et al~\cite{watts2019measuring} report of a similar
enrichment pattern and a concentration peak in the SAS of $\approx$0.5
mmol/L occuring after 10 to 15 hours. Our in-silico enrichment
patterns agree with these observations, though with an earlier
\discuss{$\sim$5 + plus spinal transit time)} rather than later peak
in the CSF spaces. In particular, we note that the substantial
individual variation observed clinically is comparable to the
variation in in-silico enrichment patterns associated with reduced CSF
production or reduced intracranial pulsatility. \mer{One sentence on
  tracer content comparison.} In terms of perivascular tracer
enrichment, Eide and Ringstad~\cite{eide2024functional} report of
first-time tracer appearance in the PVS of the MCA after 37.8 $\pm$
47.0 (M1), 53.1 $\pm$ 50.5 (M2) and 82.6 $\pm$ 61.5 (M3) min, thus
with just a 15.3 min delay between M2 and M1. Comparing with the our
FTAs in the MCA segments, we note that the baseline model predicts a
more-than 3 hour difference in FTA between the M1 and M2. Instead, the
high PVS flow scenario predicts a 24 min delay between these
segments. Moreover, the high PVS flow scenario predicts FTAs of
84--108 min (M1), 84--146 min (M2) which overlaps with the previous
ranges \discuss{modulo spinal transit time}. We interpret these
results as indicative of a net CSF flow in human surface PVS beyond
what would be induced by CSF production and cardiac peristalsis alone.
% M2 - M1 (mean) ~ 15.3 min

@@

% Limitations
In terms of limitations, our primary constraint modelling-wise is the
geometric representation of the vascular (and therefore also
perivascular) networks. Concretely, we account for no vessels below
the original MRI resolution limits, and thus only a part of the
cerebral surface vasculature and effectively not the cortical
vessels. In addition, and again due to lack of resolution, there is
uncertainty associated with the location of the vessel segments
relative to the CSF spaces and parenchyma, with some vessels
surrounded by both domains and some vessels completely entering the
parenchyma before returning to the SAS. Higher resolution MRI data
(T1w and e.g. time-of-flight, QSM or other modalities) in more
subjects would immediately allow for more detailed studies of this
aspect. As a result of the lack of an accurate representation of the
cortical and subcortical white matter vasculature, we have considered
a conservative transport model within the parenchyma, modelling
extracellular diffusion alone. To compensate, we have focused on
reporting detailed predictions in the CSF spaces and surface PVS after
not more than 24 hours. In particular, we note that, we observe tracer
accumulation at the perivascular network ends after \fixme{24} hours,
similar to that previously observed for
microspheres~\cite{bedussi2018paravascular, mestre2018flow}, which
here can be interpreted as an artifact of the incomplete network
creating up a barrier for tracer movement. Previously, we have found
that enrichment within the parenchyma is well-represented by $\approx
3\times$ enhanced diffusion augmented by local clearance e.g.~across
the blood-brain barrier or diffusion-convection with a complex pattern
of spatially-varying convective velocities of 1-10 \textmu
m/min~\cite{vinje2023human}. We will incorporate such transport
mechanisms within the parenchyma in future work. Finally, we also note
that we simulate the intracranial distribution of a tracer
concentration appearing at the craniocervical junction rather than as
an intrathecal injection. We therefore overestimate the amount of
tracer entering into the intracranial spaces; however, due to the
linearity of the mathematical model, it is valid to directly interpret
the in-silico concentrations relative to the total amount of tracer.

@@

permeability $\xi$ ($100\times$) \fixme{near} the leaf nodes of
$\Lambda_a$ to account for the transition from explicitly resolved

>>>>

% Tracer enrichment predictions
While molecular enrichment of the brain and surrounding CSF spaces
have been extensively studied in animal models, especially in murine
models in connection with studies of the glymphatic system, there are
fewer reports of contrast enrichment in human subjects over 0--24h. We
therefore mainly compare our in-silico predictions with the series of
papers by Eide and colleagues~\cite{ringstad2017glymphatic,
  ringstad2018brain, eide2021sleep, eide2024functional} and Watts et
al~\cite{watts2019measuring}. Ringstad et al.~\cite{ringstad2018brain}
report of tracer enrichment in a centripetal pattern, primarily in
regions near large cerebral surface arteries, peaking in the CSF
spaces between 6 and 9 hours, with parenchymal tracer content still
increasing after 24 hours, but with large individual
differences. Watts et al~\cite{watts2019measuring} report of a similar
enrichment pattern and a concentration peak in the SAS of $\approx$0.5
mmol/l (relative to a total amount of 1.0 mmol) occuring after 10 to 15 hours. Our in-silico enrichment
patterns agree with these observations, though with an earlier rather than later peak in the CSF spaces (2 hours after first tracer appearance at the craniocervical junction). In particular, we note that the substantial individual variation observed clinically is comparable to the variation in in-silico enrichment patterns associated with reduced CSF production or reduced intracranial pulsatility. We find that about 20\% of the tracer reaching the cranium enters the parenchyma, which aligns with the reported peak concentration values of $0.5\,$mmol/l in the SAS and $0.1\,$mmol/l in the parenchyma reported by Watts et al~\cite{watts2019measuring}.
In terms of perivascular tracer
enrichment, Eide and Ringstad~\cite{eide2024functional} report of
first-time tracer appearance in the PVS of the MCA after 37.8 $\pm$
47.0 (M1), 53.1 $\pm$ 50.5 (M2) and 82.6 $\pm$ 61.5 (M3) min after injection, thus with just a 15.3 min delay between M2 and M1. Comparing with our FTAs in the MCA segments, we note that the baseline model predicts a more-than 3 hour difference in FTA between the M1 and M2. Instead, the high PVS flow scenario predicts a 24 min delay between these segments.
Moreover, the high PVS flow scenario predicts FTAs of
84--108 min (M1) and 108--132 min (M2) after tracer appearance at the craniocervical junction, which overlaps with the previous ranges if one assumes a tracer travel time from the point of injection to the craniocervical junction of $13\,$min (as reported by Eide and Ringstad \cite{eide2024functional}). We interpret these results as indicative of a net CSF flow in human surface PVS beyond what would be induced by CSF production and cardiac peristalsis alone.
% M2 - M1 (mean) ~ 15.3 min

@@

% Limitations
In terms of limitations, our primary constraint modelling-wise is the
geometric representation of the vascular (and therefore also
perivascular) networks. Concretely, we account for no vessels below
the original MRI resolution limits, and thus only a part of the
cerebral surface vasculature and effectively not the cortical
vessels. In addition, and again due to lack of resolution, there is
uncertainty associated with the location of the vessel segments
relative to the CSF spaces and parenchyma, with some vessels
surrounded by both domains and some vessels completely entering the
parenchyma before returning to the SAS. Higher resolution MRI data
(T1w and e.g. time-of-flight, QSM or other modalities) in more
subjects would immediately allow for more detailed studies of this
aspect. As a result of the lack of an accurate representation of the
cortical and subcortical white matter vasculature, we have considered
a conservative transport model within the parenchyma, modelling
extracellular diffusion alone. To compensate, we have focused on
reporting detailed predictions in the CSF spaces and surface PVS after
not more than 24 hours. In particular, we note that, we observe tracer
accumulation at the perivascular network ends after 12 hours,
similar to that previously observed for
microspheres~\cite{bedussi2018paravascular, mestre2018flow}, which
here can be interpreted as an artifact of the incomplete network
creating up a barrier for tracer movement. Previously, we have found
that enrichment within the parenchyma is well-represented by $\approx
3\times$ enhanced diffusion augmented by local clearance e.g.~across
the blood-brain barrier or diffusion-convection with a complex pattern
of spatially-varying convective velocities of 1-10 \textmu
m/min~\cite{vinje2023human}. We will incorporate such transport
mechanisms within the parenchyma in future work. Finally, we also note
that we simulate the intracranial distribution of a tracer
concentration appearing at the craniocervical junction rather than as
an intrathecal injection. We therefore overestimate the amount of
tracer entering into the intracranial spaces; however, due to the
linearity of the mathematical model, it is valid to directly interpret
the in-silico concentrations relative to the total amount of tracer.

@@

permeability $\xi$ ($100\times$) at the the leaf nodes of
$\Lambda_a$, linearly decreasing over the corresponding cell, to account for the transition from explicitly resolved
